%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section on possible outcomes
\section{Conceptual Framework}
\label{sec:conceptual}

While state support for public higher education has stagnated at the same time as education costs rose, there are multiple possible ways that a university can respond.
\cite{NBERw23736} established that corresponding rises in tuition did not offset the falling state support, so that university spending (per student) fell in response to these persistent, negative state funding shocks.
There are multiple ways that these changes in finances may affect the faculty composition at public universities.

Universities hire new professors most years, either to expand their departments or to replace leaving professors.
The number of professors hired in each year is usually highest among non-tenured adjunct faculty, as these instructors are mostly hired on short-term contingent contracts, so that they are less costly to hire (or to not reinstate) in response to yearly changes in teaching needs.
Tenure-track assistant professors are hired most often hired with a four to six year contract, and formal agreement for tenure consideration at the end of this term.
Lastly, tenured professors have undergone review and have successfully secured a full-time appointment at their university with no expiration date; tenured professors may be hired from outside a university, though this position has the lowest yearly hiring rate among most universities.
It is common for universities to restrict or delay their faculty hiring in the spring term ahead of the next academic year, in response to unforeseen events, or changes in needs.
So it stands to reason that the hiring rate of new professors could be meaningfully affected by changes in university finances, and thus falls in state funding.

%Additionally, faculty salaries may be affected by changes in public university finances.
%When the university has a lower budget for its yearly hiring, it may respond by lowering the salary they offer to new hires.
%This possible effect may not be the same across each position of professor; tenured faculty are often hired away from another university, so that new tenured professor hires may be less likely to accept a lower salary offer from a public university.\footnote{
%    Faculty often consider their outside options, as in offers for employment and/or promotion from other universities.
%    See \cite{blackaby2005} for an overview of faculty outside options.
%}
%Yet salary for all the professors, not just new hires, may also be affected: multiple universities passed a university-wide pay-cut for their faculty in response to state budget cuts around the 2008 recession, for example.\footnote{
%    Indeed, Cornell University implemented hiring freezes and a nominal salary reduction for professors, in anticipation of a financial shock in early 2021.
%    The salary cut was not permanent, as the oncoming financial shock turned out to not be as serious as projected, so that the cut was returned to professors in the year 2021.
%}
%Faculty are not paid the same by position, so that there are financial consequences for a university promoting their own faculty.
%In response to tightening fiscal restraints, universities may be less likely to grant tenure to assistant professors (as tenured professors generally receive a salary raise and long term contract), or to promote tenured professors from associate to full professor.

Absent this analysis, it is not immediately clear why stagnating state support and changes in faculty composition at public universities would be causally linked.
Over the 21st Century there have been drastic changes to college selectivity, where the top universities have become more selective while the average university less selective \citep{hoxby2009changing}.
While the average public university is becoming less selective, it is possible that their most productive or research-focused faculty more often move to more selective and prestigious (and often private) institutions, so that over time public universities (on average) substitute towards contingent lecturers. 
This is one way in which the trends may be concurrent, but not causal.
This analysis shows that stagnating in state funding for public universities cause the changes in faculty composition, explaining about 40\% of the observed substitution towards contingent lecturers, so that it is possible that other systematic changes in higher education (such as changing selectivity) may explain the rest of the variation.

It is not immediately clear which effects will dominate, or whether tenured professors or lecturers will be affected more by changes in their university's funding.
Yet there is one empirical fact worth noting: the average assistant or tenured professor earns more than double that of a lecturer (see \autoref{tab:illinois-summary}).
Additionally, universities must pay the higher salary to tenure-track/tenured professors for a guaranteed number years, which is not true for lecturers who are mostly on shorter term contracts.
If a public university's primary obligation is to teach, and they must fulfil this objective with fewer and fewer resources, then they may substitute away from tenure-track and tenured professor towards contingent lecturers.
