%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Results section
\section{Results}
\label{sec:results}

\subsection{First-Stage Estimates}
\label{sec:results-firststage}
The shift-share instrument has a strong first-stage for universities' yearly state funding.
\autoref{tab:firststage-reg} show results of the first-stage regression among IPEDS data, separately with and without a control for tuition revenue per student.
\autoref{tab:firststage-reg} Panel A shows estimates of a state funding shift-share of \$1 per student on state funding per student at the university, and Panel B the effects of a 1\% increase in the state funding shift-share on state funding per student.
Column (1) shows that a state funding shift-share of \$10 per student is associated with \$11.76 less state funding per student at the university; $-10$\% state funding shift-share leads to $-9.77$\% state funding per student at the average university.
Estimates are presented for specifications with and without fixed effects.
Column (2) shows estimates of the first-stage without including fixed effects, and gives less precise estimates for the funding shift-share --- this may be due to systematic differences in universities, which are not accounted for when omitting fixed effects.
Columns (3) and (4) include the tuition revenue as a control (explained in \autoref{foot:control}) to exhibit estimates with or without this control variable.
Column (3) shows similar estimates to column (1) in both Panels A and B thanks to inclusion of fixed effects, so whether tuition revenue is a bad control does not affect empirical estimates.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects (not including tuition revenue as a control variable), and is the preferred specification that I proceed with;
all specifications (except column 4 in both panels) have a large F statistic on the excluded instrument.\footnote{
    I calculate the F statistic on the excluded instrument, while also accounting for fixed effects and two-way standard error clustering, by the methods provided in \cite{olea2013robust}.
}
I use the same empirical strategy in the analysis of Illinois faculty, and \autoref{tab:firststage-illinois} presents first-stage estimates among the IBHED data; results are similar to those calculated in IPEDS data.\footnote{
    Every IV specification presented in this paper uses a specification with a strong first-stage.
    Results with weak first-stages are omitted as they would not be reliable causal effects.
}
These results show the funding shift-share instrument strongly predicts a university's state funding in the first-stage estimation.

The first-stage is strong in both IPEDS and IBHED data, and the argument for exogeneity comes from independence of the state funding mechanism to any individual public university campus (as in \autoref{sec:empirics}).
The IV approach also assumes that shifts in state-wide funding for higher education (the instrument) affect outcomes only via affecting state funding for the university (exclusion restriction).
This assumption is fundamentally untestable, but is bolstered by the lack of relationship with university characteristics.
\autoref{tab:summary-quantiles} shows the mean of various university variables, across the instrument distribution.\footnote{
    The shift-share instrument is correlated with tuition revenues (and thus the sum of non-institutional revenues), so could be a presenting a threat to endogeneity.
    As mentioned above, including tuition revenue as a control (even though it could also be a bad control) does not meaningfully change first- and second-stage results in any specification, rendering this plausible threat to identification moot.
    See \autoref{foot:control} and \autoref{tab:firststage-reg}.
}
The university--years in the bottom 20\% of the instrument distribution have more research spending than those in the next 20\%; this is to be expected, as university spending is heavily related to state--wide shifts in funding.
On the other hand, the measures of selectivity (acceptance rate, 6--year graduation rate) are noticeably different across the distribution of the instrument.
While the shift--share instrument exhibits balance for many relevant university characteristics (except tuition revenues and selectivity), I include institution and year fixed effects throughout my analysis, effectively controlling for any possible confounding effects of differences between universities.

Do state funding cuts meaningfully affect universities?
While the shift-share performs well in predicting state funding (i.e., a strong instrument), it could be the case that the day-to-day operations of a university are not affected.
University could have responded to state funding cuts by cutting areas of spending unrelated to research or instruction, or fully off-set the cuts by charging higher tuition fees, leading to no effects on the university or faculty.
To allay this concern, I estimate the IV model on IPEDS data for areas of university spending  (\autoref{tab:expenditures-shock-reg}).
This exercise gauges the effects of state funding cuts on areas of university spending to gauge whether the first-stage is economically meaningful, and to see which areas of higher education operations are affected by state funding cuts.
Following a state funding cut, universities spend less on instruction, student services, operations, and grant aid.
Additionally, they charge higher tuition fees, though this rise is at most a quarter of the state funding cut, so does not come close to fully off-setting state funding cuts.
These results are in line with previous work showing state funding cuts (or tuition price limits) reduce student-focused spending \citep{NBERw23736}, financial aid \citep{miller2022making}, and lead to increased tuition fees \citep{bound2019public,webber2017state}.

\subsection{University Level, IPEDS}
\label{sec:results-ipeds}
State funding cuts lead to measurable changes in the number of faculty at US public universities.
\autoref{tab:facultycount-shock-reg} presents OLS and IV estimates for the effect of state funding cuts on the number of faculty at public universities, separately for each faculty position (lecturer, assistant professors and professors, total).
The results are presented in two different units: Panel A shows how a \$1,000 state funding cut per-student changed the count of faculty, while Panel B shows how a 1\% increase in state funding led to a percent change in the number of faculty per students.

An extra \$1,000 in state funding per student leads to the average university employing 6 more lecturers, while effects for the other positions are not discernible from zero (\autoref{tab:facultycount-shock-reg}, Panel A).
A 10\% state funding cut lead to a 4.4\% increase in the number of lecturers per student, and a decrease of 1.35\% and 1.37\% for assistant and professors, respectively.
A state funding cut of 10\% leads to a 0.65\% decrease in the total number of faculty per student (\autoref{tab:facultycount-shock-reg}, Panel B).
These are large substitution effects, and explain roughly 40\% of the trend in substitution towards lecturers over the same time period.
The average public university has roughly 36\% less state funding per student in 2021 than they did in 1990, while employing over double the number of lecturers and 20\% fewer professors (relative to enrolment, seen in \autoref{fig:mean-funding-fte}, \ref{fig:fte-perprof}).

Universities do not re-adjust by hiring more professors to make up for last year's funding cut, and so the substitution towards lecturers becomes persistent.
\autoref{fig:count-lp} shows LP estimates, showing the elasticity for professor count per student in year $t+k$ with respect to state funding in year $t$ (and cumulative intermediate cuts).
State funding cuts affect the number of lecturers and professors at universities up to 9 years after the initial funding cut.
A 10\% state funding cut leads to 4.2\% (2.9--5.5 in a 95 percent confidence interval) more lecturers per student the following year, and around 3.1\% (0.6--5.5 confidence interval) more lecturers nine years later.
A 10\% state funding cut leads to 1.5\% (0.6--2.2 in a 95 percent confidence interval) fewer assistant professors per student the following year, and around 1.1\% (0.2--2.0 confidence interval) fewer nine years later;
the same cut leads to 1.2\% (0.6--1.8 in a 95 percent confidence interval) fewer professors per student the following year, and around 0.9\% (0.1--1.7 confidence interval) fewer nine years later.

These effects are a substitution of professors for lecturers at public universities, as the total number of faculty per students are relatively unrelated to funding cuts, but the lecturer count increases and professor count decreases.
This agrees with a model for universities departments having fixed teaching requirements making human-resources decisions under constrained resources (e.g., \citealt{abe2015implications}), lending credibility to this type of model for understanding recent human resources decisions of public universities.
Furthermore, if a university must teach more students, but cannot afford to hire a full-time professor who teaches 2--3 classes a year, then it makes sense that they hire 2--3 additional lecturers on part-time contracts for each class as a replacement.
This provides simple intuition for why the lecturers' elasticity is 2--3 times larger than that for assistant/professors in \autoref{tab:facultycount-shock-reg}.
While no previous research causally connects funding cuts to substituting professors for lecturers, the phenomenon has been alluded to in popular media \citep{wiu2016} and university rankings (where more lecturers hurts a university's ranking, \citealt{usnews2023}).
The results provided here give the first real evidence for such popular claims.

\subsection{Illinois Faculty, IBHED}
Incumbent professors at Illinois universities are relatively unaffected by state funding cuts at their university.
\autoref{tab:faculty-shock-illinois-rolling} show estimates of a 10\% change in state funding per student on faculty salaries, rate of exiting the Illinois public university system, and rate of promotion.
The estimates are not discernible from zero in any specification, including the shift-share IV estimates, so that faculty seem to have little or no state funding cuts passed on to them for these outcomes.

It could be the case that effects on faculty do not materialise until multiple years after a funding cut, though the data do not support this possibility.
LP estimates show that professors' salaries, including salaries, are not significantly affected in years following funding cuts (\autoref{fig:salaries-illinois-lp-rolling}) --- they do not receive lower rates of pay raises, or delayed salary cuts, in response to state funding cuts.
Similarly, promotion rate and rate of exit from the Illinois public university system are also unaffected in later years (\autoref{fig:promoted-illinois-lp-rolling}, \ref{fig:exit-illinois-lp-rolling}).
Lecturers receive a 0.75\% increase in salary in the second and third year after a 10\% increase in state funding (\autoref{fig:salaries-lecturer-illinois-lp-rolling}), though they are they only faculty who saw any effect of state funding cuts in terms of salaries.

These results show that faculty --- both lecturers and professors --- are relatively unaffected by state funding cuts, in the Illinois public university system.
Illinois is a state with a large public university system, relative to student enrolment, with trends in state funding for education similar to national trends, and with both selective and non-selective universities.
If faculty are unaffected in Illinois, then it is unlikely that faculty in the rest of the nation are affected on average.
Note, however, that this analysis studies faculty who are already faculty.
The IBHED data do not measure the entire pool of academics applying for faculty jobs at Illinois universities, but only those who applied to, were accepted, and then started a job as a lecturer or professor.
These are incumbent faculty, and the data show they are unaffected by state funding cuts.
This raises a question: how can funding cuts lead to a large substitution of professors for lecturers, when incumbent faculty are not affected?

\subsection{Impacts on Faculty Hiring}
Turn-over is the primary channel that changes in faculty composition happen, where faculty exit their jobs and new faculty are hired to replace those leaving --- or no more are hired to replace those who left.
\autoref{fig:exit-illinois-lp-rolling} shows that Illinois faculty are no more likely to exit the Illinois system (either retire or take a new job elsewhere) following state funding cuts, so there is one remaining channel for the substitution of professors for lecturers: increased hiring of lecturers and reduced hiring of new professors.

To investigate reduced hiring of new professors, I first investigate faculty hiring rates among Illinois universities from the IBHED data.
To do so, I count the number of faculty new in the panel of all Illinois public university faculty for years 2011--2021, where the panel first year is 2010.
I then use this as the outcome in OLS and shift-share IV regressions, where each observation is an Illinois university--year and data for state funding come from IPEDS.\footnote{
    This is the same specification as the IPEDS analysis restricted to Illinois universities.
    The IPEDS data have yearly faculty counts, but not discern between faculty promotions, quits, and hires; the Illinois data can do so thanks to the panel structure of the data.
}
To study whether these effects hold true nation-wide, I use public on the number of professors at all US public universities for the entire decade 2011--2021, provided by \citep{wapman2022quantifying},
and perform the same analysis to see if state funding affected the number of professors hired.
See \autoref{sec:appendix-hiring} for further details on this data source.

\autoref{tab:facultyhires-illinois-reg} shows these estimates for Illinois public universities.
In general, faculty counts are relatively unrelated to state funding among Illinois universities; though the point estimates line up with analysis in IPEDS, the smaller number of observations mean that the precision of estimates are small and indistinguishable from zero, in both raw counts and in log units.
In the national-level \cite{wapman2022quantifying} data, universities with more state funding hire more professors across 2011--2021.
An extra 10\% in state funding is associated with 8.5\% more professors hired per students, as seen in \autoref{fig:hiring-correlation}.
I use these data as part of a shift-share analysis, showing that a 10\% state funding cut per student leads to 13\% fewer professor hires (relative to enrolment), and that these hires are roughly equal for gender of professor hired.\footnote{
    However, most of the variation in state funding cuts and substitution towards lecturers happens during the 1990--2009 period.
    So this analysis of 2011--2021 data is not ground-truth evidence for mechanisms over 1990--2009.
}

State funding cuts lead to lower professor hiring, and the persistent funding cuts to public universities led to lower professor hiring across the last three decades.
The channels for faculty turn-over were unaffected by state funding cuts in the analysis of Illinois faculty, while professor hiring is related to funding cuts in both correlational terms and in a shift-share analysis.
Combining these pieces of evidence means that reduced professor hiring was likely the main mechanism for how state funding cuts led to a substitution of professors for lecturers at public universities.

\subsection{Robustness Checks and Heterogeneity Analysis}
\label{sec:results-robustness}
It is possible that the identification strategy is not robust to some confounding issues or trends.
For example, selective institutions may have received funding cuts for other reasons, or states with lower performing institutions or declining populations strategically cut funding for only less selective universities.
I investigate these issues with two different analyses, and then investigate heterogeneity across rates of selectivity and in different decades.

First, I re-estimate the primary results from \autoref{sec:results-ipeds} with additional controls for university selectivity.
Specifically, I estimate \autoref{eqn:secondstage} with controls added for the base share of state funding, average acceptance rate and 6-year completion rate  in 2004--2006, tuition revenue, total enrolment, size of state enrolment, size of university enrolment relative to the rest of the state.
Measures of selectivity are only available in certain years, so I use these values across the entire panel.
This means that institution fixed effects would over-identify the system, so I estimate with state $+$ year fixed effects (not institution $+$ year fixed effects).
In this way, this model also works as an alternative specification test.
Additionally, this analysis uses the enrolment measure of full-time equivalent (FTE), and not reported enrolment as in the original analysis.
The first-stage is strong of the same order as the original analysis (\autoref{tab:firststage-robustness-checks}).
The relationship between state funding, in both raw count terms and log units, are roughly the same as for the main analysis (\autoref{tab:facultycount-robustness-checks}, \ref{tab:facultycount-rawcount-robustness-checks}), despite the multiple specification changes, additional controls, and different measures for enrolment.

As another robustness check, I employ is a falsification test, testing whether the effects for public universities would be picked up among a sample of private universities.\footnote{
    This is a robustness check similar to one performed in \citealt{NBERw27885}, as a robustness check for the same instrument.
}
Private universities do not rely on state funding in the same way that public universities do,\footnote{
    Some private universities do receive state funding, mostly funding for smaller scale operating costs and other specific projects.
}
so we would expect state funding to not affect their faculty counts.
If the state funding and faculty counts were related among private universities, this would be evidence that the IV model picks up otherwise concurrent trends in higher education, unrelated to stagnating state funding for public universities.
To impute a base share of reliance on state funding, I assign private universities the average reliance on state funding in the base period of universities in the same state with the same Barrons selectivity ranking, and then compute the shift-share instrument from this value.
\autoref{tab:facultycount-shock-private-robustness} shows that the first-stage is not strong among private universities; despite this, the IV results are still not significant for any faculty count, showing that state funding is not related to private universities' faculty counts.
The OLS columns show that imputed state funding is highly correlated with lecturer and professor counts at private universities.
The IV estimates are not distinguishable from zero, showing that OLS correlations are not causal relationships, and validating the IV approach for inference on universities' faculty counts.

To investigate whether universities of different selectivity respond differently to state funding cuts, I estimate the effect of state funding cuts on faculty counts separately among universities of different selectivity rates.
\autoref{tab:facultycount-heterogeneity} collects these results together, estimating \autoref{eqn:secondstage} separately by the \cite{barrons2009} selectivity rankings.
Barrons rank 1-2 are the most selective (including e.g., UC Berkeley and SUNY Binghamton), 3-4 medium selective (including e.g., Clemson University and University of Dallas), and rank 5 for the rest of universities.
The first-stage is strong in every group of universities, even among the small sample size of 12 universities in the most selective group, and 28 universities in the next selective group.
Faculty substitution rates are not distinguishable from zero among the most selective universities, but are positive and similar between both medium selective universities and unranked universities.
One might expect that the effect is driven entirely by non-selective and low ranked universities, as they are the most resource-constrained;
this heterogeneity analysis shows that even medium selective universities substitute towards lecturers following state funding cuts.

Last, I investigate when the substitution towards lecturers occurred.
To do so, I estimate the effect of state funding on faculty counts relative student enrolment separately for 1990--2009 and 2010--2021.
Substitution towards lecturer is weak to non existent in the latest decade, as noted in contemporary research \citep{hinrichs2022state}.
However, the rates are stronger when just looking at the years 1990--2010.
This reflects a key point noted in \autoref{sec:trends}: a large amount of the cuts, and general stagnation, in state funding for public universities occurred over in the 1990s and 2000s.
Additionally, the trend in rising lecturer count, and falling professor count, per student at public universities had been operating all throughout the 1990s and 2000s.
These trends did continue in the 2010s (especially for counts of assistant professors), but most of the substitution had already taken place by this point.
